### 2020-07-10 submission/manuscript.Rmd
Reorganization edits

---
title: ''
csl: msphere.csl
fontsize: 11pt
output:
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
  word_document: default
  html_document:
    df_print: paged
geometry: margin=1.0in
bibliography: references.bib
---

@@

@@

#Combine into a data frame and figure out percent colonized per day
percent_colonized_day <- rbind(day_1, day_2, day_3, day_4, day_5, day_6, day_7, day_8, day_9) %>% 
  group_by(day) %>% 
  mutate(total_n = sum(n)) %>% #Figure out total n per day (sum of cleared and colonized counts)
  filter(cfu_status == "colonized") %>% #Only need colonized rows to figure out percent colonized
  mutate(percent_colonized = (n/total_n) * 100)#Add new column for percent_colonized
percent_colonized_d7 <- percent_colonized_day %>% filter(day == 7) %>% 
  pull(percent_colonized)
n_day1 <- percent_colonized_day %>% filter(day == 1) %>% 
  pull(total_n)
n_day2 <- percent_colonized_day %>% filter(day == 2) %>% 
  pull(total_n)
n_day3 <- percent_colonized_day %>% filter(day == 3) %>% 
  pull(total_n)
n_day4 <- percent_colonized_day %>% filter(day == 4) %>% 
  pull(total_n)
n_day5 <- percent_colonized_day %>% filter(day == 5) %>% 
  pull(total_n)
n_day6 <- percent_colonized_day %>% filter(day == 6) %>% 
  pull(total_n)
n_day7 <- percent_colonized_day %>% filter(day == 7) %>% 
  pull(total_n)
n_day8 <- percent_colonized_day %>% filter(day == 8) %>% 
  pull(total_n)
n_day9 <- percent_colonized_day %>% filter(day == 9) %>% 
  pull(total_n)
```
## Results

**Clindamycin treatment renders all mice susceptible to *C. difficile* 630 colonization regardless of source.** To test how the microbiotas of mice from different sources impact colonization dynamics after clindamycin exposure, we utilized C57BL/6 mice from 6 different sources: two colonies from the University of Michigan that were split from each other 10 years ago (the Young and Schloss lab colonies) and four commercial source: the Jackson Laboratory, Charles River Laboratories, Taconic Biosciences, and Envigo (which was formerly Harlan). These 4 vendors were chosen because they represent commonly used vendors for CDI studies in mice [@Spinler2016; @Markey2018; @McKee2018; @Yamaguchi2020; @Stroke2018; @EtienneMesmin2017; @Quigley2019; @Mullish2019].

After a 13-day acclimation period for the mice ordered from vendors, all mice were treated with 10 mg/kg clindamycin via intraperitoneal injection and one day later challenged with 10^3^ *C. difficile* 630 spores (Fig. 1A). Clindamycin was chosen because we have previously demonstrated mice are rendered susceptible, but consistently cleared the CDI within 9 days [@Jenior2018; @Tomkovich2019], clindamycin is frequently implicated with human CDIs [@Guh2018], and is also part of the antibiotic treatment for the frequently cited 2008 CDI mouse model [@Chen2008]. The day after infection, *C. difficile* was detectable in all mice at a similar level (median CFU range: `r format(min_median_d1, digits = 2, scientific=TRUE)`-`r format(max_median_d1, scientific=TRUE, digits = 2)`; *P*~FDR~ = `r padj_d1_cfu`), indicating clindamycin rendered all mice susceptible regardless of source (Fig. 1B). Interestingly, variation in *C. difficile* CFU levels across sources of mice emerged from days 3-7 post-infection (all *P*~FDR~ $\le$ `r max_padj_d3to7`; Fig. 1B and Table S1), suggesting mouse source is associated with *C. difficile* clearance. We conducted two experiments approximately 3 months apart and while the colonization dynamics were similar across most sources of mice, there was some variation between the 2 experiments, particularly for the Schloss and Envigo mice (Fig. S1A-B). Although *C. difficile* 630 causes mild symptoms in mice comparied to other *C. difficile* strains [@Theriot2011], we also saw that weight change significantly varied across sources of mice with the most weight lost two days post-infection (Fig. 1C and Table S2). Importantly, there was also one Jackson and one Envigo mouse that died between 1- and 3-days post-infection during the second experiment. Interestingly, mice obtained from Jackson, Taconic, and Envigo tended to lose more weight (although there was variation between experiments with Schloss and Envigo mice), have higher *C. difficile* CFU levels and take longer to clear the infection compared to the other sources of mice, which was particularly evident 7 days post-infection (Fig. 1B-C, Fig. S2C-D), when `r format(percent_colonized_d7, digits = 1)`% of the mice were still colonized with *C. difficile* (Fig. S1E). By 9 days post-infection the majority of the mice from all sources had cleared *C. difficile* (Fig. S1C) with the exception of 1 Taconic mouse from the first experiment and 2 Envigo mice from the second experiment. Thus, clindamycin rendered all mice susceptible to *C. difficile* 630 colonization, regardless of source, but there was significant variation in disease phenotype across the sources of mice.

```{r alpha}
richness_stats_all <- read_tsv("../data/process/richness_stats_all_days.tsv")
padj_dn1_richness <- richness_stats_all %>% filter(day == -1) %>% pull(p.value.adj)

shannon_stats_all <- read_tsv("../data/process/shannon_stats_all_days.tsv")
padj_dn1_shannon <- shannon_stats_all %>% filter(day == -1) %>% pull(p.value.adj)
```
**Bacterial communities consistently vary across sources despite antibiotic and infection perturbations.** Given the well known variation in mouse microbiomes across breeding colonies [@Alegre2019], we hypothesized that the variation in *C. difficile* clearance could be explained by microbiota variation across the 6 sources of mice. We used 16S rRNA gene sequencing to characterize the fecal bacterial communities from the mice over the course of the experiment. Since antibiotics and other risk factors of CDIs are associated with decreased microbiota diversity [@Ross2016], we first examined alpha diversity measures across the 6 sources of mice. Examining the bacterial communities at baseline, prior to clindamycin treatment, there was a significant difference in the number of observed OTUs (*P*~FDR~ = `r padj_dn1_richness`), but not Shannon diversity index (*P*~FDR~ = `r padj_dn1_shannon`) across sources of mice (Fig. 2A-B and Table S3). As expected, the clindamycin treatment decreased richness and Shannon diversity across all sources of mice, and richness and Shannon diversity started to increase 1 day post-infection (Fig. 2C-D). Interestingly, significant differences in diversity metrics across sources (*P*~FDR~ < 0.05) emerged after both clindamycin and *C. difficile* infection, with Charles River mice having higher richness and Shannon Diversity than most of the other sources (Fig 2C-F and Table S4). Although the Charles River mice had more diverse microbiotas, the Young and Schloss lab mice were also able to clear *C. difficile* faster than the other sources, suggesting microbiota diversity alone does not explain the observed variation in *C. difficile* colonization across vendors.
```{r adonis}
adonis_stats_all <- read_tsv("../data/process/adonis_all.tsv")
r_sq_ordered_adonis <- adonis_stats_all %>% arrange(desc(r_sq))
rsq_adonis_source <- r_sq_ordered_adonis %>% pull(r_sq) %>% max
p_adonis_source <- r_sq_ordered_adonis %>% pull(p) %>% max

source_adonis_dn1 <- read_tsv("../data/process/adonis_dn1_source.tsv")
adonis_schloss_rsq_sum <- source_adonis_dn1 %>% filter(source == "Schloss") %>% select(r_sq) %>% sum
p_adonis_schloss_max <- source_adonis_dn1 %>% filter(source == "Schloss") %>% select(p) %>% max
adonis_young_rsq_sum <- source_adonis_dn1 %>% filter(source == "Young") %>% select(r_sq) %>% sum
p_adonis_young_max <- source_adonis_dn1 %>% filter(source == "Young") %>% select(p) %>% max
```
Next, we compared the community structure of mice from the 6 sources over the course of the experiment using principal coordinate analysis (PCoA) of the $\theta_{YC}$ distances. Permutational multivariate analysis of variance (PERMANOVA) analysis revealed source was the dominant factor that explained the observed variation across fecal communities (R^2^ = `r rsq_adonis_source`, *P* = `r p_adonis_source`) followed by interactions between cage (mice from the same sources were housed together, primarily at a density of 2 mice per cage) and day of the experiment (Movie S1 and Table S5). Mice that are co-housed have been shown to have similar gut microbiotas due to coprophagy [@Nguyen2015] and since mice within the same source were housed together, it is not surprising that cage also contributes to the observed microbiota variation. Since the majority of the perturbations happened over the initial days of the experiment, we decided to focus on the communities at baseline (day -1), after clindamycin treatment (day 0), and post-infection (day 1). For all 3 timepoints, the source and cage interaction significantly explained most of the observed community variation (combined R^2^ = `r dn1_rsq_sum`, `r d0_rsq_sum`, `r d1_rsq_sum`, respectively; *P* = `r p_adonis_intial`; Fig. 3 and Table S6). We also compared baseline communities across the 2 experiments, and found experiment and cage significantly explained the observed variation only for the Schloss and Young lab mouse colonies (Fig. S2A-B and Table S7). However, most of the vendors also clustered by experiment (Fig. S2C-D, F), suggesting there was some community variation between the 2 experiments within each source. Thus, source was the factor that explained the most variation observed in the bacterial communities. Importantly with the exception of the Schloss and Young colonies, the community of each source clustered apart from one another suggesting each community had a unique response to clindamycin treatment and *C. difficile* challenge.

Since there was some variation in the microbiota structure of mice from the same source between experiments at baseline, we next looked at how similar the communities were within the same source and between sources in response to clindamycin treatment and *C. difficile* challenge (Fig. 4) . The baseline communities varied most between experiments for Schloss, Young, and Envigo mice and variation between sources of mice was high (Fig. 4A). Clindamycin treatment reduced the variation between experiments within Schloss, Young and Jackson mice and some of the variation between sources diminished, particularly for the Schloss, Young, and Charles River mice (Fig. 4B). Post-infection, the community variation started to increase within sources of mice and variation between sources of mice started to increase towards previous levels (Fig. 4C). By using mice from multiple sources we were able to increase the number of microbiota communities we tested with the clindamycin *C. difficile* colonization mouse model. 
```{r source_taxa}
#Number of different otus across sources at day -1, 0, and 1 timepoints
source_otus <- read_tsv("../data/process/otu_stats_dn1to1_combined.tsv")
source_otus_dn1 <- source_otus %>% filter(day == -1) %>% nrow()
source_otus_d0 <- source_otus %>% filter(day == 0) %>% nrow()
source_otus_d1 <- source_otus %>% filter(day == 1) %>% nrow()

```
After finding differences at the community level, we next identified the bacteria that varied across sources of mice over the initial days of the experiment. We examined bacterial relative abundances at the operational taxonomic unit (OTU) level. Focusing on the baseline communities first, there were `r source_otus_dn1` OTUs (Table S8) with relative abundances that varied across sources. Clindamycin treatment reduced the number of taxa with relative abundances that varied across sources to `r source_otus_d0` OTUs (Table S8). After *C. difficile* challenge, there were `r source_otus_d1` OTUs (Table S8) with significantly different relative abundances across sources, as the communities started to recover from antibiotic treatment. In spite of the experimental perturbations that occurred during these 3 timepoints, there were `r shared_source_otus_dn1to1` OTUs (Fig 5A-C) with relative abundances that consistently varied across sources. Importantly, some of the OTus that consistently varied across sources also shifted with clindamycin treatment. For example, *Proteus* increased after clindamycin treatment, but only in Taconic mice. *Enterococcus* was primarily found only in mice purchased from commercial vendors and also increased after clindamycin treatment. In summary, mouse bacterial communities varied significantly according to source throughout the course of the experiment and a consistent subset of bacterial taxa remained different across sources regardless of clindamycin and *C. difficile* challenge.

```{r clindamycin_taxa}
clindamycin_otus <- read_tsv("../data/process/otu_stats_dn1to0.tsv")
clindamycin_otus_sig <- clindamycin_otus %>% filter(p.value.adj < 0.05) %>% nrow()
```
**Clindamycin treatment alters a subset of taxa that were found in all sources.** Although there were bacteria that consistently varied across sources, we also wanted to identify the bacteria that shifted after clindamycin treatment, regardless of source. By analyzing all mice that had sequence data from fecal samples collected at baseline and after clindamycin treatment, we identified `r clindamycin_otus_sig` OTUs that were altered after clindamycin treatment (Fig. 6 and Tables S9). Interestingly, when we compared the list of significant clindamycin impacted bacteria with the bacteria that consistently varied across groups over the initial 3 timepoints of our experiment, we found 3 OTUs (*Lachnospiraceae* (OTU 130), *Lactobacillus* (OTU 6), *Enterococcus* (OTU 23)) overlapped (Fig. 5, Fig. 6C-D). These findings demonstrate that clindamycin has a consistent impact on the fecal bacterial communities of mice from all sources and only a subset of the OTUs also varied across sources.

@@

#Venn diagram calculations for OTUs----
venn_otus_source <- length(dayn1to1_and_interp_combined)
venn_otus_intersect <- length(intersect(dayn1to1_and_interp_combined, paired_and_interp_combined))
venn_otus_clind <-  length(paired_and_interp_combined)
```
**Source-specific and clindamycin impacted bacteria distinguish *C. difficile* colonization status in mice.** After identifying taxa that varied by source, changed after clindamycin treatment, or both, we next wanted to determine which taxa were influencing the variation in *C. difficile* colonization at day 7 (Fig. 1D, Fig. S1C). We trained three L2-regularized logistic regression models with input bacterial community data from the baseline, post-clindamycin, and post-infection timepoints of the experiment to predict *C. difficile* colonization status on day 7 (Fig. S3A-B). All models were better at predicting *C. difficile* colonization status on day 7 than random chance (all *P* $\le$ `r format(max_p_vs_random, scientific = TRUE, digits = 2)`; Table S12). Interestingly, the model based on the post-clindamycin (day 0) community OTU data performed significantly better than all other models with an area under the receiving operator characteristic curve (AUROC) of `r median_otu_day0_AUROC` (*P*~FDR~ $\le$ `r format(max_padj_otu_day0_comp, scientific = TRUE, digits = 2)` for pairwise comparisons; Table S13). Thus, we were able to use bacterial relative abundance data alone to differentiate mice that had cleared *C. difficile* before day 7 from the mice still colonized with *C. difficile* at that timepoint. Interestingly, the model built with OTU relative abundance data post-clindamycin treatment had the best performance, suggesting how the bacterial community responds to clindamycin treatment has the greatest influence on subsequent *C. difficile* colonization dynamics.

Next, to examine the bacteria that were driving each model’s performance, we pulled out the top 20 taxa that had the highest absolute feature weights in each of the 6 models (Tables S14-15). First, we looked at OTUs from the model with the best performance that was based on the post-clindamycin treatment bacterial community data. While most of the 20 OTUs had low relative abundances on day 0, *Enterobacteriace*, *Bacteroides* and *Proteus* had high relative abundances in at least one source of mice and significantly varied across sources (Fig. 7A). Next, the top 20 taxa from each model were compared to the list of taxa that varied across source at the same timepoint (Fig. 5 and Tables S8-9) and the taxa that were altered by clindamycin treatment (Fig. 6 and Table S10-11). We found a subset of OTUs that were important to the model and overlapped with bacteria that varied by either source, clindamycin treatment, or both (Fig. S4, S5A-C). Combining the overall results for the 3 OTU models identified `r venn_otus_source` OTUs associated with source, `r venn_otus_clind` OTUs associated with clindamycin treatment, and `r venn_otus_intersect` OTUs associated with both (Fig. 7B). Several OTUs (*`r list_key_otus`*) appeared across at least 2 models, so we examined how the relative abundances of these key taxa varied over the course of the experiment (Fig. 8). Throughout the experiment, there was at least 1 timepoint where relative abundances of these OTUs significantly varied across sources (Table S13). Interestingly, there were no OTUs that emerged as consistently enriched or depleted in mice that were colonized past 7 days post-infection with *C. difficile* 630, suggesting multiple bacteria influence the time needed to clear the infection. Together, these results suggest the initial bacterial communities and their responses to clindamycin influence the clearance of *C. difficile*.

@@

By running our CDI model with mice from 6 different sources, we were able to identify bacterial taxa that were unique to sources throughout the experiment as well as taxa that were universally impacted by clindamycin. We trained L2 logistic regression models with baseline, post-clindamycin treatment, and post-infection fecal community data that could predict whether mice cleared *C. difficile* by 7 days post-infection better than random chance. We identified *`r list_key_otus`* (Fig. 8) as candidate bacteria within these communities that were influencing variation in *C. difficile* colonization dynamics since these bacteria were all important in the logistic regression models and varied by source, were impacted by clindamycin treatment, or both. Overall, our results demonstrate clindamycin is sufficient to render mice from multiple sources susceptible to CDI and only a subset of the interindividual microbiota variation across mice from different sources was associated with the time needed to clear *C. difficile*.

@@

**(vii) Statistical analysis.** All statistical tests were performed in R (v `r paste(version$major, ".", version$minor, sep = "")`) [@r_citation_2018]. The Kruskal-Wallis test was used to analyze differences in *C. difficile* CFU, mouse weight change, and alpha diversity across vendors with a Benjamini-Hochberg correction for testing multiple timepoints, followed by pairwise Wilcoxon comparisons with Benjamini-Hochberg correction. For taxonomic analysis and generation of logistic regression model input data, *C. difficile* (OTU 20) was removed. Bacterial relative abundances that varied across sources at the OTU level were identified with the Kruskal-Wallis test with Benjamini-Hochberg correction for testing all identified OTUs, followed by pairwise Wilcoxon comparisons with Benjamini-Hochberg correction. OTUs impacted by clindamycin treatment were identified using the Wilcoxon signed rank test with matched pairs of mice samples for day -1 and day 0. To determine whether classification models had better performance (test AUROCs) than random chance (0.5), we used the one-sample Wilcoxon signed rank test. To examine whether there was an overall difference in predictive performance across the 6 classification models we used the Kruskal-Wallis test followed by pairwise Wilcoxan comparisons with Benjamini-Hochberg correction for multiple hypothesis testing. The tidyverse package was used to wrangle and graph data (v `r paste(packageVersion("tidyverse")[1])`) [@Tidyverse2019].

>>>>

---
title: ''
csl: msphere.csl
fontsize: 11pt
output:
  word_document: default
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
  html_document:
    df_print: paged
geometry: margin=1.0in
bibliography: references.bib
---

@@

## Results
```{r alpha}
richness_stats_all <- read_tsv("../data/process/richness_stats_all_days.tsv")
padj_dn1_richness <- richness_stats_all %>% filter(day == -1) %>% pull(p.value.adj)

shannon_stats_all <- read_tsv("../data/process/shannon_stats_all_days.tsv")
padj_dn1_shannon <- shannon_stats_all %>% filter(day == -1) %>% pull(p.value.adj)
```

```{r adonis}
adonis_stats_all <- read_tsv("../data/process/adonis_all.tsv")
r_sq_ordered_adonis <- adonis_stats_all %>% arrange(desc(r_sq))
rsq_adonis_source <- r_sq_ordered_adonis %>% pull(r_sq) %>% max
p_adonis_source <- r_sq_ordered_adonis %>% pull(p) %>% max

source_adonis_dn1 <- read_tsv("../data/process/adonis_dn1_source.tsv")
adonis_schloss_rsq_sum <- source_adonis_dn1 %>% filter(source == "Schloss") %>% select(r_sq) %>% sum
p_adonis_schloss_max <- source_adonis_dn1 %>% filter(source == "Schloss") %>% select(p) %>% max
adonis_young_rsq_sum <- source_adonis_dn1 %>% filter(source == "Young") %>% select(r_sq) %>% sum
p_adonis_young_max <- source_adonis_dn1 %>% filter(source == "Young") %>% select(p) %>% max
```

```{r source_taxa}
#Number of different otus across sources at day -1, 0, and 1 timepoints
source_otus <- read_tsv("../data/process/otu_stats_dn1to1_combined.tsv")
source_otus_dn1 <- source_otus %>% filter(day == -1) %>% nrow()
source_otus_d0 <- source_otus %>% filter(day == 0) %>% nrow()
source_otus_d1 <- source_otus %>% filter(day == 1) %>% nrow()

```

**Microbiota variation is high between mice from different sources.** To test how the microbiotas of mice from different sources impact colonization dynamics after clindamycin exposure, we utilized C57BL/6 mice from 6 different sources: two colonies from the University of Michigan that were split from each other 10 years ago (the Young and Schloss lab colonies) and four commercial source: the Jackson Laboratory, Charles River Laboratories, Taconic Biosciences, and Envigo (which was formerly Harlan). These 4 vendors were chosen because they represent commonly used vendors for CDI studies in mice [@Spinler2016; @Markey2018; @McKee2018; @Yamaguchi2020; @Stroke2018; @EtienneMesmin2017; @Quigley2019; @Mullish2019]. We used 16S rRNA gene sequencing to characterize the fecal bacterial communities across sources at baseline.

Since antibiotics and other risk factors of CDIs are associated with decreased microbiota diversity [@Ross2016], we first examined alpha diversity across the 6 sources of mice. Intially, there was a significant difference in the number of observed operational taxonomic units (OTUs), but not Shannon diversity index (*P*~FDR~ = `r padj_dn1_richness` and *P*~FDR~ = `r padj_dn1_shannon`, respectively) across sources of mice (Fig. 1A-B and Table S1). Next, we compared the community structure of mice from the 6 sources over the course of the experiment using principal coordinate analysis (PCoA) of the $\theta_{YC}$ distances. Permutational multivariate analysis of variance (PERMANOVA) analysis revealed source and cage interactions explained most of the observed variation across fecal communities explained most of the observed community variation (combined R^2^ = `r dn1_rsq_sum`, *P* = `r p_adonis_intial`, Fig. 3 and Table S2). Mice that are co-housed have been shown to have similar gut microbiotas due to coprophagy [@Nguyen2015] and since mice within the same source were housed together, it is not surprising that cage also contributed to the observed microbiota variation.  Since we conducted two experiments approximately 3 months apart, we also compared baseline communities across the 2 experiments. Experiment and cage significantly explained the observed variation only for the Schloss and Young lab mouse colonies (Fig. S2A-B and Table S2). However, most of the vendors also clustered by experiment (Fig. S2C-D, F), suggesting there was some community variation between the 2 experiments within each source, particularly for Schloss, Young, and Envigo mice (Fig. S2G-H). After finding differences at the community level, we next identified the bacteria that initially varied across sources of mice. There were `r source_otus_dn1` OTUs with relative abundances that varied across sources (Fig. 1D and Table S3). By using mice from six sources we were able to increase the number of microbiota communities to test with the clindamycin *C. difficile* colonization mouse model.

@@

#Combine into a data frame and figure out percent colonized per day
percent_colonized_day <- rbind(day_1, day_2, day_3, day_4, day_5, day_6, day_7, day_8, day_9) %>% 
  group_by(day) %>% 
  mutate(total_n = sum(n)) %>% #Figure out total n per day (sum of cleared and colonized counts)
  filter(cfu_status == "colonized") %>% #Only need colonized rows to figure out percent colonized
  mutate(percent_colonized = (n/total_n) * 100)#Add new column for percent_colonized
percent_colonized_d7 <- percent_colonized_day %>% filter(day == 7) %>% 
  pull(percent_colonized)
n_day1 <- percent_colonized_day %>% filter(day == 1) %>% 
  pull(total_n)
n_day2 <- percent_colonized_day %>% filter(day == 2) %>% 
  pull(total_n)
n_day3 <- percent_colonized_day %>% filter(day == 3) %>% 
  pull(total_n)
n_day4 <- percent_colonized_day %>% filter(day == 4) %>% 
  pull(total_n)
n_day5 <- percent_colonized_day %>% filter(day == 5) %>% 
  pull(total_n)
n_day6 <- percent_colonized_day %>% filter(day == 6) %>% 
  pull(total_n)
n_day7 <- percent_colonized_day %>% filter(day == 7) %>% 
  pull(total_n)
n_day8 <- percent_colonized_day %>% filter(day == 8) %>% 
  pull(total_n)
n_day9 <- percent_colonized_day %>% filter(day == 9) %>% 
  pull(total_n)
```
**Clindamycin treatment renders all mice susceptible to *C. difficile* 630 colonization, but clearance time varies across sources.** After the acclimation period, all mice were treated with 10 mg/kg clindamycin via intraperitoneal injection and one day later challenged with 10^3^ *C. difficile* 630 spores (Fig. 2A). Clindamycin was chosen because we have previously demonstrated mice are rendered susceptible, but consistently cleared *C. difficile* within 9 days [@Jenior2018; @Tomkovich2019], clindamycin is frequently implicated with human CDIs [@Guh2018], and is also part of the antibiotic treatment for the frequently cited 2008 CDI mouse model [@Chen2008]. The day after infection, *C. difficile* was detectable in all mice at a similar level (median CFU range: `r format(min_median_d1, digits = 2, scientific=TRUE)`-`r format(max_median_d1, scientific=TRUE, digits = 2)`; *P*~FDR~ = `r padj_d1_cfu`), indicating clindamycin rendered all mice susceptible regardless of source (Fig. 2B). Interestingly, variation in *C. difficile* CFU levels across sources of mice emerged from days 3-7 post-infection (all *P*~FDR~ $\le$ `r max_padj_d3to7`; Fig. 2B and Table S4), suggesting mouse source is associated with *C. difficile* clearance. We conducted two experiments and while the colonization dynamics were similar across most sources of mice, there was some variation between the 2 experiments, particularly for the Schloss and Envigo mice (Fig. S2A-B). Although *C. difficile* 630 causes mild symptoms in mice comparied to other *C. difficile* strains [@Theriot2011], we also saw that weight change significantly varied across sources of mice with the most weight lost two days post-infection (Fig. 2C and Table S5). Importantly, there was also one Jackson and one Envigo mouse that died between 1- and 3-days post-infection during the second experiment. Interestingly, mice obtained from Jackson, Taconic, and Envigo tended to lose more weight, have higher *C. difficile* CFU levels and take longer to clear the infection compared to the other sources of mice (although there was variation between experiments with Schloss and Envigo mice), which was particularly evident 7 days post-infection (Fig. 2B-C, Fig. S2C-D), when `r format(percent_colonized_d7, digits = 1)`% of the mice were still colonized with *C. difficile* (Fig. S2E). By 9 days post-infection the majority of the mice from all sources had cleared *C. difficile* (Fig. 2C) with the exception of 1 Taconic mouse from the first experiment and 2 Envigo mice from the second experiment. Thus, clindamycin rendered all mice susceptible to *C. difficile* 630 colonization, regardless of source, but there was significant variation in disease phenotype across the sources of mice.

```{r clindamycin_taxa}
clindamycin_otus <- read_tsv("../data/process/otu_stats_dn1to0.tsv")
clindamycin_otus_sig <- clindamycin_otus %>% filter(p.value.adj < 0.05) %>% nrow()
```
**Clindamycin treatment alters bacteria in all sources, but a subset of bacterial differences across sources persists.** Given the well known variation in mouse microbiomes across breeding colonies [@Alegre2019], we hypothesized that variation in *C. difficile* clearance would be explained by microbiota variation across the 6 sources of mice. We used 16S rRNA sequencing to characterize microbiota variation across sources after clindamycin treatment and throughout the rest of the experiment. PERMANOVA analysis revealed source was the dominant factor that explained the observed variation across fecal communities (R^2^ = `r rsq_adonis_source`, *P* = `r p_adonis_source`) followed by interactions between cage (mice from the same sources were housed together, primarily at a density of 2 mice per cage) and day of the experiment (Movie S1 and Table S2). Since the majority of the perturbations happened over the initial days of the experiment, we decided to focus on the communities after clindamycin treatment (day 0) and post-infection (day 1).

As expected, clindamycin treatment decreased richness and Shannon diversity across all sources of mice (Fig. 3A-B). Interestingly, significant differences in diversity metrics across sources (*P*~FDR~ < 0.05) emerged after clindamycin treatment, with Charles River mice having higher richness and Shannon diversity than most of the other sources (Fig 3A-B and Table S1). The variation in community structures was also diminished by clindamycin treatment (Fig. 3C and Table S2). Next we identified the bacteria that shifted after clindamycin treatment, regardless of source by analyzing all mice that had sequence data from fecal samples collected at baseline and after clindamycin treatment. We identified `r clindamycin_otus_sig` OTUs that were altered after clindamycin treatment (Fig. 3D and Table S6) in most mice. Despite the significant impact of clindamycin, there were still `r source_otus_d0` OTUs (Fig. 3E and Table S3) with relative abundances that significantly varied across sources. Interestingly, when we compared the list of significant clindamycin impacted bacteria with the bacteria that varied across sources post-clindamycin, we found 4 OTUs (*Enterobacteriaceae (OTU 1)*, *Lachnospiraceae* (OTU 130), *Lactobacillus* (OTU 6), *Enterococcus* (OTU 23)) overlapped (Fig. 3D-E and Tables S3, S6). These findings demonstrate that clindamycin has a consistent impact on the fecal bacterial communities of mice from all sources and only a subset of the OTUs continued to vary across sources.

**Microbiota variation across sources is maintained after *C. difficile* challenge.** Post-infection, significant differences in diversity metrics  were maintained across sources (*P*~FDR~ < 0.05, Fig 4A-B and Table S1). Although the Charles River mice had more diverse microbiotas, the Young and Schloss lab mice were also able to clear *C. difficile* faster than some of the other sources, suggesting microbiota diversity alone does not explain the observed variation in *C. difficile* colonization across sources. Source and cage interaction continued to explain most of the observed community variation (combined R^2^ = `r d1_rsq_sum`, respectively; *P* = `r p_adonis_intial`; Fig. 4D and Table S2). After *C. difficile* challenge, there were `r source_otus_d1` OTUs (Fig. 4D and Table S3) with significantly different relative abundances across sources. In spite of the experimental perturbations that occurred, there were `r shared_source_otus_dn1to1` OTUs with relative abundances that consistently varied across sources at baseline, post-clindamycin, and post-infection. Importantly, some of the OTUs that consistently varied across sources also shifted with clindamycin treatment. For example, *Proteus* increased after clindamycin treatment, but only in Taconic mice. *Enterococcus* was primarily found only in mice purchased from commercial vendors and also increased after clindamycin treatment. In summary, mouse bacterial communities varied significantly according to source throughout the course of the experiment and a consistent subset of bacterial taxa remained different across sources regardless of clindamycin and *C. difficile* challenge.

@@

#Venn diagram calculations for OTUs----
venn_otus_source <- length(dayn1to1_and_interp_combined)
venn_otus_intersect <- length(intersect(dayn1to1_and_interp_combined, paired_and_interp_combined))
venn_otus_clind <-  length(paired_and_interp_combined)
```
**Baseline, post-clindamycin, and post-infection community data can predict mice that will clear *C. difficile* by 7 days post-infection.** After identifying taxa that varied by source, changed after clindamycin treatment, or both, we next wanted to determine which taxa were influencing the variation in *C. difficile* colonization at day 7 (Fig. 2D, Fig. S2C). We trained three L2-regularized logistic regression models with input bacterial community data from the baseline, post-clindamycin, and post-infection timepoints of the experiment to predict *C. difficile* colonization status on day 7 (Fig. S3A-B). All models were better at predicting *C. difficile* colonization status on day 7 than random chance (all *P* $\le$ `r format(max_p_vs_random, scientific = TRUE, digits = 2)`; Table S7). Interestingly, the model based on the post-clindamycin (day 0) community OTU data performed significantly better than the other models with an area under the receiving operator characteristic curve (AUROC) of `r median_otu_day0_AUROC` (*P*~FDR~ $\le$ `r format(max_padj_otu_day0_comp, scientific = TRUE, digits = 2)` for pairwise comparisons; Table S7). Thus, we were able to use bacterial relative abundance data alone to differentiate mice that had cleared *C. difficile* before day 7 from the mice still colonized with *C. difficile* at that timepoint. Interestingly, the model built with OTU relative abundance data post-clindamycin treatment had the best performance, suggesting how the bacterial community responds to clindamycin treatment has the greatest influence on subsequent *C. difficile* colonization dynamics.

Next, to examine the bacteria that were driving each model’s performance, we pulled out the top 20 OTUs that had the highest absolute feature weights in each of the 6 models (Table S8). First, we looked at OTUs from the model with the best performance that was based on the post-clindamycin treatment bacterial community data. While most of the 20 OTUs had low relative abundances on day 0, *Enterobacteriace*, *Bacteroides* and *Proteus* had high relative abundances in at least one source of mice and significantly varied across sources (Fig. 5A). Next, the top 20 taxa from each model were compared to the list of taxa that varied across source at the same timepoint (Fig. 1D, 3E, 4D and Table S3) and the taxa that were altered by clindamycin treatment (Fig. 3D and Table S6). We found a subset of OTUs that were important to the model and overlapped with bacteria that varied by either source, clindamycin treatment, or both (Fig. S4). Combining the overall results for the 3 models identified `r venn_otus_source` OTUs associated with source, `r venn_otus_clind` OTUs associated with clindamycin treatment, and `r venn_otus_intersect` OTUs associated with both (Fig. 5B). Several OTUs (*`r list_key_otus`*) appeared across at least 2 models, so we examined how the relative abundances of these OTUs varied over the course of the experiment (Fig. 6). Throughout the experiment, there was at least 1 timepoint where relative abundances of these OTUs significantly varied across sources (Table S9). Interestingly, there were no OTUs that emerged as consistently enriched or depleted in mice that were colonized past 7 days post-infection with *C. difficile* 630, suggesting multiple bacteria influence the time needed to clear the infection. Together, these results suggest the initial bacterial communities and their responses to clindamycin influence the clearance of *C. difficile*.

@@

By running our CDI model with mice from 6 different sources, we were able to identify bacterial taxa that were unique to sources throughout the experiment as well as taxa that were universally impacted by clindamycin. We trained L2 logistic regression models with baseline, post-clindamycin treatment, and post-infection fecal community data that could predict whether mice cleared *C. difficile* by 7 days post-infection better than random chance. We identified *`r list_key_otus`* (Fig. 6) as candidate bacteria within these communities that were influencing variation in *C. difficile* colonization dynamics since these bacteria were all important in the logistic regression models and varied by source, were impacted by clindamycin treatment, or both. Overall, our results demonstrate clindamycin is sufficient to render mice from multiple sources susceptible to CDI and only a subset of the interindividual microbiota variation across mice from different sources was associated with the time needed to clear *C. difficile*.

@@

**(vii) Statistical analysis.** All statistical tests were performed in R (v `r paste(version$major, ".", version$minor, sep = "")`) [@r_citation_2018]. The Kruskal-Wallis test was used to analyze differences in *C. difficile* CFU, mouse weight change, and alpha diversity across sources with a Benjamini-Hochberg correction for testing multiple timepoints, followed by pairwise Wilcoxon comparisons with Benjamini-Hochberg correction. For taxonomic analysis and generation of logistic regression model input data, *C. difficile* (OTU 20) was removed. Bacterial relative abundances that varied across sources at the OTU level were identified with the Kruskal-Wallis test with Benjamini-Hochberg correction for testing all identified OTUs, followed by pairwise Wilcoxon comparisons with Benjamini-Hochberg correction. OTUs impacted by clindamycin treatment were identified using the Wilcoxon signed rank test with matched pairs of mice samples for day -1 and day 0. To determine whether classification models had better performance (test AUROCs) than random chance (0.5), we used the one-sample Wilcoxon signed rank test. To examine whether there was an overall difference in predictive performance across the 6 classification models we used the Kruskal-Wallis test followed by pairwise Wilcoxan comparisons with Benjamini-Hochberg correction for multiple hypothesis testing. The tidyverse package was used to wrangle and graph data (v `r paste(packageVersion("tidyverse")[1])`) [@Tidyverse2019].
